---
title: "ISYE 6501"
subtitle: "Homework 2"
author: "Artur Cabral, Marta Bras, Pedro Pinto, Katie Price"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    df_print: paged
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(kernlab)
library(kknn)
library(caret)
library(e1071)
library(knitr)
library(kableExtra)
library(colorspace)
```

# Question 3.1

***Using the same data set (credit_card_data.txt or credit_card_data-headers.txt) as in Question 2.2, use the ksvm or kknn function to find a good classifier:***
  
##	1. Cross-validation: kkn model
  
### Importing and preparing the data 
  
```{r}
credit_card <- read.table("credit_card_data-headers.txt", sep= "", header = TRUE)
credit_card[,11]  <- factor(credit_card[,11])

```


### Choosing best model {.tabset}

The kknn package has a function - train kknn - that performs crossvalidation in 2 ways: 1) k-fold crossvalidation (cv.kknn) and 2) leave-one-out crossvalidation (train.kknn). The caret package also offers a train function to perform k-fold crossvalidation. 

In the next steps, we perform the crossvalidation with the 3 functions, to understand which one and what number of neighbors provide the highest accuracy.

#### cv.kknn

cv.kkn uses k-fold crossvalidation: the dataset is split into a K number of sections/folds where each fold is used as a testing set at some point.

```{r}
# setting seed for reproducibility
set.seed(123) 

# looping through different numbers of neighbors to find the perfect number using cv.kknn for crossvalidation
v_accuracy <- vector()

for (k in 1:30){
  pred <- c()
  knn <- cv.kknn(R1 ~ .,credit_card , k = k, kcv = 10,  scale=TRUE)
  pred <- c(pred, as.integer(knn[[1]][,2])-1) # rounding prediction to either one or zero
  accuracy <- sum(pred == credit_card[,11]) / nrow(credit_card)
  cat("\n For K = ", k, "accuracy = ", accuracy, "\n")
  v_accuracy <- c(v_accuracy,accuracy)
}

#best accuracy
cat("\n The best value of k is K = ", match(c(max(v_accuracy,na.rm = TRUE)),v_accuracy), " with accuracy = ", max(v_accuracy,na.rm = TRUE), "\n")

#outputting the results 
number <-  seq((1: length(v_accuracy)))
accuracy <- cbind(round(v_accuracy,4), number)
colnames(accuracy) <- c("accuracy","k")
accuracy <- as.data.frame(accuracy)

ggplot(accuracy, aes(x=k, y=accuracy)) +
  geom_point(color = "#159957") +
  ggtitle("Accuracy levels for different k values")
```


#### train.kknn

train.kkn uses leave-one-out crossvalidation - the data is trained on every point except one, for which the prediction is made. 

```{r}
# setting seed for reproducibility
set.seed(123) 

# Fitting model with multiple k values
model <- train.kknn(R1~.,credit_card,kmax=30,scale=TRUE)
best_parameter = as.integer(model$best.parameters[2]$k)
min_error = model$MISCLASS[best_parameter]
accuracy = (1 - min_error) 


cat("\n The best value of k is K = ", best_parameter, " with accuracy = ",  accuracy)

#outputting the results 
plot(model)
```


```{r}
#trying different kernells
model_2 <- train.kknn(R1~., credit_card, kmax = 30, kernel = 
                        c("triangular", "rectangular", "epanechnikov", "optimal"), distance = 2)

plot(model_2)
```


#### train from caret package

We can also use the caret package which has the train function to perform k-fold crossvalidation.

```{r}
#setting seed for reproducibility
set.seed(123) 

#creating the train control parameter wiht 10 folds and 5 repetitions
trControl <- trainControl(method  = "repeatedcv",
                          number  = 10,
                          repeats = 5)

#ftting the model with the defined train control
model <- train(R1 ~ A1 + A2 + A3 + A8 + A9 + A10 + A11 + A12 + A14 + A15,
               method     = "knn",
               trControl  = trControl,
               metric     = "Accuracy",
               data       = credit_card,
               preProcess = c("center", "scale"),
               tuneLength = 30)

#best parameters
index = which.max(model$results[,2])
accuracy = model$results[index,2]
k = model$results[index,1]

cat("\n The best value of k is K = ", k, " with accuracy = ",  accuracy)


#outputting the results 
plot(model)
```


#### Conclusion

The best accuracy is obtained with the **17 neighbors** using **cv.kknn**.

## 2. Cross-validation: ksvm model {.tabset}

### k-fold Cross-Validation:
```{r}
#Performing k-fold cross-validation usign the tune function from the e1071 library:

#Preparing data

credit_card_svm <- data.matrix(credit_card)
class <- credit_card_svm[,11]
predictors <-  credit_card_svm[,1:10]

#creating dataframe
credit_card_2 <- data.frame(predictors =predictors, class)

# find optimal cost of misclassification
svmfit <- svm(class~., data = credit_card_2, kernel = "linear", cost = 10, gamma = 1)

tune.out <- tune(svm, class~., data = credit_card_2, kernel = "linear",
                 ranges = list(cost = c(0.001, 100,0.01, 0.1, 1, 5, 10, 100)))

# extract the best model
(bestmod <- tune.out$best.model)

#graph of error values for different C values

plot(tune.out)
```

### KSVM model and results

Using the 0.01 value for C from tune function, we get that the best model is:

```{r}
best_model <- ksvm(predictors, class, kernel="vanilladot", type = "C-svc", C=0.01,scaled=TRUE)

#printing parameters
best_model 

#predicting for dataset
ypred <- predict(best_model, predictors)

#plotting confusion matrix
misclass <- table(predict = ypred, truth = class)
misclass

```

Our model has an accuracy of 86.4% in the training dataset. It produces 286 true negatives and 279 true positives, vs 17 false positives and 72 false negatives. 


### Decision boundary function

```{r}
# calculate a1…am
a <- colSums(best_model@xmatrix[[1]] * best_model@coef[[1]])
a
# calculate a0
a0 <- best_model@b
a0

```

The decision boundary function is: 

*-0.08198854 + -0.0001500738A1 + -0.0014818294A2 + 0.0014083130A3 + 0.0072863886A8 + 0.9916470037A9 +  -0.0044661236A10 + 0.0071482899A11 + -0.0005468386A2 + -0.0016930578A14 + 0.1054824270A15*





## 3. Training, validation, and testing: kknn


### Splitting data into training, validation and testing

```{r}
#setting seed for reproducibility
set.seed(123)

#setting 60% for training
ind <- sample(1:nrow(credit_card),floor(nrow(credit_card)*0.60)) 
credit_card_training <- credit_card[ind,]

#setting half of remaining for validation
credit_card_remaining <- credit_card[-ind,]
ind_2 <- sample(1:nrow(credit_card_remaining),floor(nrow(credit_card_remaining)*0.50)) 
credit_card_val <- credit_card_remaining[ind_2,]

#setting half of remaining for testing
credit_card_test <- credit_card_remaining[-ind_2,]
```



### Building the model and checking accuracy 


```{r}

# looping through different numbers of neighbors to find the perfect number using kknn function

v_accuracy <- vector()

for (k in 1:30){
  pred <- c()
  knn <- kknn(R1 ~ .,credit_card_training,credit_card_val, k = k,  scale=TRUE)
  pred <- c(pred, as.integer(fitted(knn))-1) # rounding prediction to either one or zero
  #pred <- pred-1
  accuracy <- sum(pred == credit_card_val[,11]) / nrow(credit_card_val)
  cat("\n For K = ", k, "accuracy = ", accuracy, "\n")
  v_accuracy <- c(v_accuracy,accuracy)
}

#best accuracy in validation dataset

cat("\n The best value of k is K = ", match(c(max(v_accuracy,na.rm = TRUE)),v_accuracy), " with accuracy in validation dataset of = ", max(v_accuracy,na.rm = TRUE), "\n")
```

### Checking accuracy on the testing dataset 

```{r}
# fitting model with best k number
knn_test <- kknn(R1 ~.,credit_card_training,credit_card_test, k = match(c(max(v_accuracy,na.rm = TRUE)), v_accuracy), scale=TRUE)

#predict in test dataset
pred_test <- as.integer(fitted(knn_test))-1

#accuracy in testing dataset
accuracy_test <- sum(pred_test == credit_card_test[,11]) / nrow(credit_card_test)


cat("\n The performance of the model in the test dataset is: ", accuracy_test, "\n")

```


# Question 4.1

***Describe a situation or problem from your job, everyday life, current events, etc., for which a clustering model would be appropriate. List some (up to 5) predictors that you might use.***

Clustering would be appropriate to study earth-quakes. Based on the areas hit by an earthquake in a region, clustering can help analyse the next probable location where the earthquake can occur.   
The predictors would be for example:   

1. Past Events: previous earthquakes occurences in the same region, and its parameters 
2. Changes in velocity of primary and secondary seismic waves
3. Random emissions
4. Electromagnetic anomalies


Identifying regions at-risk might help governments better alocate resources to these areas to develop better infrastructure and prevent catastrophic outcomes. 
  
# Question 4.2
  
***Use the R function kmeans to cluster the points as well as possible. Report the best combination of predictors, your suggested value of k, and how well your best clustering predicts flower type.***
  
  
## Importing and brief EDA of the data 

```{r}
#importing data
iris <- iris
classifiers <- iris[,1:4]
class <- iris[,5]
```
```{r}
#getting a glimpse of data
glimpse(iris)
```


```{r}
#number of observations per plant type
dt <- iris %>% group_by(Species) %>% 
  summarize(number_plants = length(Species)
  )

kable(dt, caption = "Number of observations per specie type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))

#main statistics for sepal length
dt <- iris %>% group_by(Species) %>% 
  summarize(
    average = mean(Sepal.Length),
    max = max(Sepal.Length),
    min = min(Sepal.Length),
    var = round(var(Sepal.Length),2))

kable(dt, caption = "Main statistics for sepal length") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))

#main statistics for sepal width
dt <- iris %>% group_by(Species) %>% 
  summarize(
    average = mean(Sepal.Width),
    max = max(Sepal.Width),
    min = min(Sepal.Width),
    var = round(var(Sepal.Width),2))

kable(dt, caption = "Main statistics for sepal width") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))

#main statistics for petal length
dt <- iris %>% group_by(Species) %>% 
  summarize(
    average = mean(Petal.Length),
    max = max(Petal.Length),
    min = min(Petal.Length),
    var = round(var(Petal.Length),2))

kable(dt, caption = "Main statistics for petal length") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))

#main statistics for petal width
dt <- iris %>% group_by(Species) %>% 
  summarize(
    average = mean(Petal.Width),
    max = max(Petal.Width),
    min = min(Petal.Width),
    var = round(var(Petal.Width),2))

kable(dt, caption = "Main statistics for petal width") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))
```


```{r}
#scatter plot 1
scatter <- ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) 
scatter + geom_point(aes(color=Species, shape=Species)) +
  xlab("Sepal Length") +  ylab("Sepal Width") +
  ggtitle("Sepal Length and Width per species")
```

```{r}
#scatter plot 2
scatter <- ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) 
scatter + geom_point(aes(color=Species, shape=Species)) +
  xlab("Petal Length") +  ylab("Petal Width") +
  ggtitle("Petal Length and Width per species")
```

## Finding optimal number of clusters

Because the starting assignments are random, we can specify nstart = 20. The algorithm will try 20 different random starting assignments and  select the one that has a lowest within cluster variation. 

For the value of k, we know 3 is a good starting point but we can try different values to check which one provides the best classification model.

We can also start by scaling the data, since that is a good practice for kmean.

```{r}
#scaling data
set.seed(123)

scale_data = matrix(nrow=nrow(classifiers), ncol=4)

for (i in (1 : length(classifiers))) {
  scale_data[,i] <- (classifiers[,i]-min(classifiers[,i]))/(max(classifiers)-min(classifiers[,i]))
}
```

```{r}
#trying different k values in the scaled data
k_max = 10
cluster = matrix(nrow=nrow(classifiers), ncol=k_max)
distance = matrix(nrow=k_max, ncol=length(classifiers))
total_withincluster = matrix(nrow=1, ncol=k_max)

for (i in (1 : k_max)) {
  models <- kmeans(scale_data[,1:4], i, nstart = 20)
  cluster[,i] <- models$cluster
  distance[i,] <- models$centers[models$cluster[i],]
  total_withincluster[,i] <- models$tot.withinss
}
```

```{r}
#creating a Within Cluster plot
plot(1:k_max,total_withincluster, type= "b", xlab = "Number of clusters(k)", ylab = "Within cluster sum of squares")
```

Looking at the Within Cluster graph, there seems to be an elbow at k = 3. We can use k = 3 in the final cluster model.

## Building the model with 3 clusters for scaled data

```{r}
#model with k = 3
model <- kmeans(scale_data[,1:4], 3, nstart = 20)

model
```


```{r}
#plotting the classifications for k = 3 with cluster vector
table(cluster[,3],class)
```

Looking at the results, setosa appears to be class 2, versicolor class 1 and virginica class 3. Setosa was always correctly classified while versicolor was incorrectly classified as virginica in 9 cases. Virginica was incorrectly classified as setosa in 15 cases.
Accuracy would be given by: (50+41+35)/(50+41+9+15+35) = 84%.
  
```{r}
#plotting clusters for petal length and petal width
ggplot(iris, aes(Petal.Length, Petal.Width, color = as.factor(cluster[,3]), shape=Species)) + geom_point() + ggtitle("3 clusters with Petal Width and Petal Length")

#plotting clusters for sepal width and petal width
ggplot(iris, aes(Petal.Width, Sepal.Width, color = as.factor(cluster[,3]), shape=Species)) + geom_point() + ggtitle("3 clusters with Sepal Width and Petal Width")

```
  

## Building the model with 4 clusters for scaled data

We can also see how the model with 4 clusters performed and how the model with 3 clusters performed in the variable before being centered.

```{r}
table(cluster[,4],class)
```

```{r}
#plotting clusters for petal length and petal width
ggplot(iris, aes(Petal.Length, Petal.Width, color = as.factor(cluster[,4]), shape=Species)) + geom_point() + ggtitle("4 clusters with Petal Width and Petal Length")
```

The model using 4 clusters, classifies setosa as class 3, versicolor as class 1 and class 3 and virginica as class 3. The accuracy for this mdoel is given by (50+27+23+33)/(50+27+23+33+14+3) = 87%. The accuracy is higher for k = 4.

## Trying with original data

```{r}
#not scaled data k = 3
model <- kmeans(iris[,1:4], 3, nstart = 20)
model
```

```{r}
#not scaled data k = 3
model <- kmeans(iris[,1:4], 4, nstart = 20)
model
```

The model using **4 clusters** and **original data** provides the highest accuracy.
